---
title: "Univariate Statistics"
format: html
editor: source
author:  Simon J. Kiss
execute:
  echo: true
  message: false
  warning: false
toc: true
toc-location: left
---

```{r }
#| label: output-lines
#| echo: false
library(knitr)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options))  # pass to default hook
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines)==1) {        # first n lines
    if (length(x) > lines) {
      # truncate the output, but add ....
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(more, x[lines], more)
  }
  # paste these lines together
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
})
``` 

## Univariate Statistics

Here we look at some basic statistical ways to assess *one variable*, e.g. *univariate statistics in R. 

```{r}
#| label: load-libraries 

library(haven)
library(here)
library(tidyverse)

```
## Import the data

Let's work with the Survey of Household Spending dataset that is stored in the `course_data` subfolder. Note that that folder is stored in a folder *above* the current working directory. If you have opened up the `05_univariate_statistics.Rproj` folder then you should be in that working directory.

For me, on my computer, that directory is: 

```{r}
#| label: working-directory
getwd()

```

But the data set is stored in a folder *next to* `05_Univariate_Statistics`. The `here()` package can help us get there. 

By adding `..` to `here()` we move up one directory, then we can add the name of the directory `course_data`.

```{r}
#| label: data-import
library(haven)
shs<-read_sav(here('..','course_data', 'shs.sav'))


```

Let's play with these two variables. Note, that in the `select()` command you can change the names the same way as with the `rename()` command.
```{r}
shs %>% 
  select(Income=HH_TotInc,Dwelling=DwelTyp, Taxes=TX001)->shs_subset
#Convert labelled variables to factors

shs_subset<-as_factor(shs_subset)
```


## Measures of Central Tendency

One of the first things you want to know with a variable is a measure of where the center is, or what the typical value of that variable is. These are variables of central tendency. There are three: mean, median and mode. All three can be used to describe numeric variables, but only mode can be used to describe categorical variables. 

### Mean

We find the mean of a variable with `mean()`

```{r}
#| label: mean
mean(shs_subset$Income)
```

Note that if there are any missing values in a variable, you will need to set `na.rm=T`. 

```{r}
#| label: mean-na-rm
mean(shs_subset$Income, na.rm=T)
```

You can always save the result.

```{r}
#| label: save-mean
mean_income<-mean(shs_subset$Income)
mean_income
```

### Median

The trouble with the average is that it is vulnerable to outliers. The classic joke is that if Bill Gates walks into a room with 100 people suddenly the average income of the room skyrockets. In cases like this, the better measure is the `median()` which finds the value at which half the values are below and half are above. 

```{r}
#| label: median


median(shs_subset$Income)

```

### Mode

In fact, there is no built-in command to take the mode of a variable. The way to do it is to run a `table()` command on a variable and count the categories that have the most responses. 

```{r}
#| label: table

table(shs_subset$Dwelling)
```