---
title: "Bivariate Statistics"
format: html
editor: source
toc: true
toc-location: left
execute:
  warning: false
  message: false
  echo: true

---


```{r }
#| label: output-lines
#| echo: false
library(knitr)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options))  # pass to default hook
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines)==1) {        # first n lines
    if (length(x) > lines) {
      # truncate the output, but add ....
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(more, x[lines], more)
  }
  # paste these lines together
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
})
``` 

## Bivariate Statistics or Measurements of Association

Univariate statistics (e.g. mean, median, mode, standard deviation) describe individual variables. We often want to describe the *relationship* between two variables. 

To understand associations between variables it is *essential* to understand *levels of measurement*. This is a concept that is central to statistics and is implemented in R in ways that we have already met. 
### Levels of measurement

In R, if you recall, variables can be of different types: `numeric`, `character` or `factors`. In statistics, the two largest categories of levels of measurements are: numeric and categorical variables. 

Numeric variables can be *continuous* or *discrete*. Continuous variables can be subdivided into infinity. Examples include money (e.g. $50.555672828 makes sense), temperature (e.g. 24.234575828 C). Discrete variables are also numeric but cannot be meaningfully subdivided into decimals. For example, we might measure how many children a family has. this variable can only ever be integers (e.g. 1,2,3,4 or 0).

Categorical variables cannot be measured in terms of numbers. Examples of this might include a person's sex (male or female), their field of study (political science or economics) or their province of residence. 

::: {tip.callout}
Just because provinces are clearly categorical variables, does not stop us from *storing* them as numbers. The trick is conceptual. Saskatchewan is just different from Alberta, it is not one more than Alberta in the way that $10 is $1 greater than $9. 

In R categorical variables are stored as factors meaning that each response is stored as a number and then a level is mapped to each number. It just saves on memory to do so. 

:::

Measuring the association between two variables depends on what level of measurement you are dealing with. 

## Import the data

Let's import the `shs` dataset from the `course_data` subfolder.
```{r}
#| label: load-libraries

# tidyverse for lots of stuff
library(tidyverse)
# here for getting working directories
library(here)
# haven for importing datasets
library(haven)
#labelled to look through large datasets
library(labelled)
#crosstable package for crosstabs
library(crosstable)
```

```{r}
#| label: import-data
#import data
shs<-read_sav(here("../course_data/shs.sav"))
```

```{r}
#| label: select-columns

lookfor(shs, "province")
shs %>% 
  #Select income, taxes, dwelling type and housing tenure
  select(Income=HH_TotInc, Taxes=TX001,Dwelling=DwelTyp, Tenure=Tenure, Province=Prov)->shs_subset
#glimpse
glimpse(shs_subset)


```

Note that two of our variables are these `chr+lbl`. These we need to convert to factors. We can do thise by putting the whole data frame in `as_factor`

```{r}
#| label: asfactor
shs_subset<-as_factor(shs_subset)
#Compare now
glimpse(shs_subset)
```

## Numeric - Numeric

To test whether there is a relationship between two numeric variables, we can use the most basic form of *correlation* the Pearson correlation coefficient. 

It takes an x and a y variable as its arguments. To access these, requires the `data_frame$variable_name` syntax.

```{r}
#| label: pearson

cor(shs_subset$Income, shs_subset$Taxes)

```

A different type of correlation is the Spearman correlation which is meant for ordinal variables (think for example the level of education a person has completed or the job title a person has). These are categories but there is an order to them. 

The spearman's correlation coefficient is also useful for data that is *not normally distributed*.

This code produces two histograms that shows that our two variables are *not normally distributed*.

```{r}
#| label: income-taxes-histogram

shs_subset %>% 
  pivot_longer(Income:Taxes) %>% 
  ggplot(., aes(x=value))+geom_histogram()+facet_grid(~name)
```

So we can specify using the Spearman correlation coefficient in this way.

```{r}
#| label: spearman
cor(shs_subset$Income, shs_subset$Taxes, method=c("spearman"))
```

So, no surprise here, household income is *highly* correlated with the amount of taxes people pay. 

## Categorical - Categorical

### Crosstabulations
We can't use the same statistics to measure the strength of association between two categorical variables. Instead, we use the chi-squared statistic and Cramer's V. 

We start with a cross-tab of two categorical variables. The most basic command is `table()`

Let's test whether or not there is a relationship between the dwelling type and the tenure of residence (e.g. owned or rented)
```{r}
#| label: crosstab

table(shs_subset$Dwelling, shs_subset$Tenure)

```
A few things: notice that the variable that is first specified in the `table()` argument is showed in the rows. The table that is specified second is visible in the columns. Second, notice that these are *frequencies* i.e. just counts. On the surface of it, we can sort of see that there is a relationship, i.e. that most people wwho live in apartments also rent. But without calculating the percentages it is hard to tell.

Sometimes reversing the rows and columns can also be informative. When the crosstabulation is created this way we are (I think ) implicitly hypothesizing that the type of home a person lives in dictates the financing. But maybe it is actually the reverse. A person can only afford to rent and so the end up in an apartment rather than a single-detached home.

We can just reverse the arguments. At the same time, I can also show you the `with()` argument. It loads the data frame for the following arguments and makes it a bit quicker to run. 


```{r}
#| label: crosstab2
with(shs_subset, table(Tenure, Dwelling))
```

Notice that this gives us a little bit more information: We also get the variable names printed and that can be helpful!

To get a table of percentages, we use the command `prop.table()`. 

Sometimes, we will want to save the `table()` output in an object and then reuse that. 

```{r}
#| label: tbl-prop-table
#| tbl-cap: Row percentages of housing tenure and dwelling type. 
#Save the table
tab1<-with(shs_subset, table(Tenure, Dwelling))
#print
tab1
#Wrap in prop.table
prop.table(tab1, 1)
```

Notice the `1` in `prop.table()` It specifies to calculate the proportions *by row*. Sometimes you will want column percentages in which case you provide a `2`. And not providing a `1` or a `2` provides the corner percentages, meaning the percentage of the whole table that is in each cell. 

In @tbl-prop-table we get 57.1% of people who rent living in an apartment with 77% of people who own with a mortgage living in a single, detached home. That certainly seems like there is a relationship. But it's not a formal test. 

We can do that with the chi-squared test. It takes a table object as its argument
```{r}
#| label: chi-sq

chisq.test(tab1)

```

In this output we get the following information: 
- the chi-squared statistic is 4651.6
- the degrees of freedom (df) is 4
- the p-value (the probability that a table completely generated at random would generate this variation in cell values) is extremely low, less then 0.00000000000000002. 

So, it is extremely unlikely a table like this would appear by chance alone.

But that does not tell us whether there is a *strong* association or not. For this, we can test Cramer's V. 
It is in the `rcompanion` package, please install this if necessary. 
```{r}
#| label: cramer-v

#install.packages("rcompanion")
library(rcompanion)
cramerV(tab1)
```
This output gives is a Cramer's V of 0.43, which is a moderate correlation. 

### Exporting Tables

Exporting crosstabulations of the type that we have been dealing with here is actually a little bit tricky. A lot of times, we want to have a table that produces *both* the raw counts from a `table()` command *and* the percentages from `prop.table()` but there is no ready-made built-in way to produce that in R. The easiest way to get at that stuff is with the `tabyl` function in the `janitor` package. 

```{r}
#| label: load-crosstable
#install.packages('janitor')
library(janitor)


```


Check the help page for the basic syntax.

```{r}
#| label: help-crosstable
?tabyl
```

We provide the `data` and then the columns that we want to tabulate. 

```{r}
#| label: crosstable1
tabyl(shs_subset, Tenure)

```

Notice that we get both the counts of each category *and* the proportions (in decimal format). We can convert those to percentages with some `adorn_*` functions. 


```{r}
#| label: crosstable-percent
tabyl(shs_subset, Tenure) %>% 
  adorn_pct_formatting()


```



The best way to export this output is to use the `flextable()` package. We pipe the output of `tabyl)` to `as_flextable()`

Multiple formats are available:
```{r}
#| label: export-flextable

#install.package('flextable')
library(flextable)
#Save the flextable as crosstab1
tabyl(shs_subset, Tenure) %>% 
  adorn_pct_formatting() %>% 
  as_flextable()->tabyl1
#Save crosstab1 as an html file
save_as_html(tabyl1, path=here("table1.html"))
# save as word document 
save_as_docx(tabyl1, path="table1.docx")

```

We can also specify two variables to come up with a crosstabulation of two variables. 

```{r}
#| label: crosstab-tab1

tabyl(shs_subset, Dwelling, Tenure)

```

It's not that hard to get the percentages and the counts in one table. 

```{r}
#| label: adorn


tabyl(shs_subset, Dwelling, Tenure) %>% 
  adorn_percentages()

```
By default those return the row percentages, but we can get the column percentages by using the `denominator` argument.

```{r}
#| label: adorn-columns


tabyl1<-tabyl(shs_subset, Dwelling, Tenure) %>% 
  adorn_percentages(denominator="col")

#print
tabyl1
```

Then we can format those percentages to be a bit neater *and* add the n's back in. Note, that this *must* happe in a particular order. First, the percentages must be formatted, then the n's added back in. 
```{r}
#| label: adorn-n


tabyl1<-tabyl(shs_subset, Dwelling, Tenure) %>% 
  #Add the column percentages
  adorn_percentages(denominator="col") %>% 
  #Format the percentages neatly
  adorn_pct_formatting %>% 
  #Add the ns back in 
  adorn_ns()
#print 
tabyl1
```

WE can also add row and column totals for extra infromation

```{r}
#| label: adorn-totals


tabyl1<-tabyl(shs_subset, Dwelling, Tenure) %>% 
  #Add the totals
  adorn_totals("both") %>% 
  #Add the column percentages
  adorn_percentages(denominator="col") %>% 
  #Format the percentages neatly
  adorn_pct_formatting() %>% 
    #Add the ns back in 
  adorn_ns() 
#print
tabyl1

```

`tabyl()` can also create a three-way table. Let's see if we can the relationship between housing dwelling and housing tenure changes by province. Do make this simpler, let's just compare Ontario and Quebec.

```{r}
#| label: three-way-table 
tabyl2<-shs_subset %>% 
  #Just filter the provinces of Ontario and Quebec
  filter(Province=="Ontario"|Province=="Quebec") %>% 
  # Make the tabyl with Dwelling, Tenure and province
  tabyl(., Dwelling, Tenure, Province) %>% 
  #Add the percentages
  adorn_percentages() %>% 
  #format the percentages
  adorn_pct_formatting %>%
  #add the ns
  adorn_ns()
tabyl2
```


Exporting a `tabyl` is a lot easier than exporting a `table()` or `prop.table()`. It's one of the major reasons I use it. There are three leading packages that export output from R into something useable: `knitr` and `kableExtra` combined, `gt` and `flextable`. I'm going to suggest we use `flextable`. 
First we have to convert the tabyl we want to save to a `flextable` object with `as_flextable` then we can use a series of `save_as_*` functions to save it out. 

WE can save it as an html file, a docx or a powerpoint. Or even an image like a jpeg. 
```{r}
#| label: flextable

#install.packages('flextable')
library(flextable)
tabyl1<-tabyl1 %>% 
  #Convert to flextable
  as_flextable()
#save as html
save_as_html(tabyl1, path=here("tabyl1.html"))
#save as png
save_as_image(tabyl1, path=here("tabyl1.png"))
#save as powerpoint
save_as_pptx(tabyl1, path=here("tabyl1.pptx"))
#save as docx
save_as_docx(tabyl1, path=here("tabyl1.docx"))
```



## Categorical - Numeric

There is no real test of association between a categorical and a numeric variable. Instead, we normally test for differences in means between the groups of the categorical variable. 

When we have a categorical variable with *two* categories as an independent variable and a *numeric* variable as the dependent variable, we can use the *t-test* for the differences in group means. 

In our `shs_subset` dataset we have the variable `Tenure` which has three categories:
```{r}
#| label: tbl-tenure
#| tbl-cap: "Responses to the question of housing tenure."
table(shs_subset$Tenure)

```

Let's reduce this to two categories so that we can use it for the t-test.

```{r}
#| label: recode-tenure
library(car)
shs_subset %>% 
  mutate(Tenure2=Recode(Tenure, "'Owned with mortgage'='Owned'; 'Owned without mortgage'='Owned'", levels=c("Owned", "Rented")))->shs_subset
#Check
table(shs_subset$Tenure2)

```

Now we can compare the average income of owners with the average income of renters using the *t-test*. This is the formula. 
$$ t= \frac{\overline{x_1}-\overline{x_2}}{\sqrt{\frac{s^2_1}{n_1}+\frac{s^2_2}{n_2}}} $$ 

If we look at the top of the fraction we see a term that subtracts the average of group 2 from the average of group 1, so we just subtract the mean of group one from the mean of the other. 

But then we divide that difference in average by something. Specifically, underneath the square root we have this:

$$ \frac{s^2_1}{n_1}+\frac{s^2_2}{n_2}$$

These compare the *variance* (which is the square root of the standard deviation) of group 1 divided by the sample size of group 1 and we add the *variance* of group 2 divided by the sample size of group 2. This gives us a ratio of the difference in group means divided by how much overall dispersion there is in the *whole sample*. 

The number that is spit out a the end is a $$t$$-statistic. This number comes from a statistical distribution of variables that was discovered by a William Gossett, who worked at the Guiness Brewery in Dublin. He noticed that taking *random samples* from a population resulted in a distribution of values that were *almost, but not quite,* normal. Where the normal distribution the $$t$$ distribution is *slightly* different, but when the sample size is > 30, the differences are miniscule. The following graph shows the distribution of $$ t $$ scores. 

One other piece of information that is necessary: the distribution of t-scores depends on the sample size, which is expressed in the term *degrees of freedom*. Larger samples have larger degrees of freedom, smaller samples have smaller degrees of freedom. 



In a completely random process, with the number of degrees of freedom that match what our example will be, t-scores are distributed as in @fig-t-distribution.


```{r echo=F}
#| label: fig-t-distribution
#| fig-cap: Distribution of t-scores from a random number generatig process.
ggplot(data=data.frame(x=seq(-3,3,1)), aes(x))+
  stat_function(fun=dt, geom="line", args=list(df=10471))+
  stat_function(fun=dt, geom="area", fill="lightgrey", xlim=c(-1,1),args=list(df=10471))+
   stat_function(fun=dt, geom="area", fill="darkgrey", xlim=c(-1,-2),args=list(df=10471))+
     stat_function(fun=dt, geom="area", fill="darkgrey", xlim=c(1,2),args=list(df=10471))+
       stat_function(fun=dt, geom="area", fill="black", xlim=c(2,4),args=list(df=10471))+
         stat_function(fun=dt, geom="area", fill="black", xlim=c(-2,-4),args=list(df=10471))+
  labs(x="T-Score", y="Probability")+theme_bw()+scale_x_continuous(breaks=seq(-3,3,1))




```


Notice that t-scores higher than 3 or below -3 are *extremely* unlikely in a completely random process. This is what we use to test the hypotheses that group means in a sample are generated completely randomly or are the product of a real process. Effectively the logic runs like this: The probability of getting a t-score above 3 in a universe where there was no relationship between my independent and my dependent variable is *very low*; it would be in the tail region above 3. Therefore I conclude that it was *not* a random generation process and that in fact there *is* a relationship between my independent and my dependent variable.

Now, to run the t-test, we simply use the command `t.test(dependent_variable ~ independent_variable, data=data_frame)`
Note that in R the `~` represents a `formula` object. What is on the left-hand side is the *dependent variable* and what is on the right-hand side is the independent variable. 

In this case, our dependent variable is the numeric variable (`Income`) and the *independent* variable is the categorical variable, `Tenure2`. So what we are doing is testing the hypothesis that there is *no* difference in average income between owners and renters. 

So the syntax looks like this: 

```{r}
#| label: actual-t-test

t.test(Income~Tenure2, data=shs_subset)

```

We get a lot of information here, let's go through it. First, we get a t-score of 47.288 on the 10471 degrees of freedom. We can actually draw the line of where that t-score is on the distribution above in @fig-t-distribution. 

```{r echo=F}
#| label: fig-t-distribution-2
#| fig-cap: Distribution of t-scores from a random number generatig process.
ggplot(data=data.frame(x=seq(-50,50,1)), aes(x))+
  stat_function(fun=dt, geom="line", args=list(df=10471))+
  stat_function(fun=dt, geom="area", fill="lightgrey", xlim=c(-1,1),args=list(df=10471))+
   stat_function(fun=dt, geom="area", fill="darkgrey", xlim=c(-1,-2),args=list(df=10471))+
     stat_function(fun=dt, geom="area", fill="darkgrey", xlim=c(1,2),args=list(df=10471))+
       stat_function(fun=dt, geom="area", fill="black", xlim=c(2,4),args=list(df=10471))+
         stat_function(fun=dt, geom="area", fill="black", xlim=c(-2,-4),args=list(df=10471))+
  labs(x="T-Score", y="Probability")+theme_bw()+scale_x_continuous(breaks=seq(-50,50,1))+
  geom_vline(xintercept=47.288, col="darkred", linewidth=2, linetype=2)




```

See where that red line is? *A whole series of t-scores generated from random samples of a world where there were no relationship between tenure and household income would be distributed like in the graph below. And red line indicates where this particular sample lands. You can see it is very improbable that you would get a sample like this in a world where there was no relationship*. Therefore we conclude that there *is* in fact a relationship. 

And that makes perfect sense.  

## Exercises 

::: {note.callout}
1. Use the `lookfor()` command to find the variable that represents `spousal income`  `respondent income` and `sex` and `dwelling tenure`
2. Use the `select()` command to select only those four variables. Rename each variable in the `select()` command with useful and informative names. Avoid whitespace.
3. Save the dataframe of these four variables into a smaller dataframe with a name of your choice.
4. Correlate `respondent_income` with `spousal_income`.
5. Create a cross-tab of `sex` and `housing_tenure` using the `table()` command.
6. Perform a chisq-test to test whether or not there is a difference in whether female respondents and male respondents report different housing tenures.
7. Conduct a t-test of whether or not females and males report different respondent incomes. 
 
:::

